Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-30 arxiv/examples.tex
revise

\begin{corollary}[Derived from Theorem \ref{thm_many_tasks}]\label{prop_same_cov}
	In the setting described above, the total prediction loss of the $t$ tasks satisfies that
	\begin{align*} 
	 \sum_{i=1}^t L(\hat{\beta}_i^{\MTL})&= (t-r) \kappa^2 \cdot \frac{\tr\Sigma}{p}  + \sigma^2 \cdot \frac{r}{\rho-1} +\OO\left( p^{-c_\infty} \cdot\frac{\rho t(\kappa^2 + d^2) \|\Sigma\|}{(\sqrt{\rho}-1)^2 \kappa^2  \|\Sigma\| + \sigma^2}   \left( t \kappa^2 \|\Sigma \| + \frac{\sigma^2 r}{\rho-1} \right)  \right) .
	 %& = \sum_{i=r+1}^t \lambda_i^2 (\Sigma^{1/2} B^{\star}) + \sigma^2 \frac{r}{\rho-1} +\OO\left(  p^{-c_\infty} \frac{t(\kappa^2 + d^2)\cdot (p^{-1}\tr \Sigma)}{(\sqrt{\rho}-1)^2 \kappa^2 \cdot (p^{-1}\tr \Sigma) + \sigma^2}   \left( t \kappa^2 \frac{\tr \Sigma}{p} + \frac{\sigma^2 r}{\rho-1} \right) \right) \\ 
%	 &= (t-r) \kappa^2 \cdot \frac{\tr\Sigma}{p}  + \sigma^2 \cdot \frac{r}{\rho-1} \\
%	 &+\OO\left( p^{-c_\infty}t(\kappa^2+d^2) \cdot p^{-1/2} \|\Sigma\|_F + p^{-c_\infty} \frac{t(\kappa^2 + d^2)\cdot (p^{-1}\tr \Sigma)}{(\sqrt{\rho}-1)^2 \kappa^2 \cdot (p^{-1}\tr \Sigma) + \sigma^2}   \left( t \kappa^2 \frac{\tr \Sigma}{p} + \frac{\sigma^2 r}{\rho-1} \right)  \right) .
	\end{align*}
%and $\lambda_{r+1} (\Sigma^{1/2} B^{\star}) \ge \cdots\ge  \lambda_{t} (\Sigma^{1/2} B^{\star})$ denote the smallest $(t-r)$ singular values of $\Sigma^{1/2} B^{\star}$.
\end{corollary}
On the other hand, the total STL predication loss is 
\begin{align*}
 \sum_{i=1}^t L(\hat{\beta}_i^{\STL}) =\sigma^2\cdot \frac{t}{ \rho-1}  +\OO(p^{-c_\infty}\cdot t \sigma^2 ).
	 \end{align*}

@@

\begin{corollary}[Derived from Theorem \ref{thm_main_RMT}]\label{cor_MTL_loss}
	In the setting described above, assume that there exists a small constant $c_0>0$ such that the signal-to-noise ratios satisfy that
	\be\label{choiceofpara0}
		p^{ c_0} \le \frac{p\kappa^2}{\sigma^2 }  \le p^{1-c_0},\quad  \frac{pd^2}{\sigma^2}=\OO(1).
	\ee
	%Let $B, A_1, A_2$ be the global minimum of equation \eqref{eq_mtl}. Then
	The prediction loss of the hard parameter sharing estimator for task two satisfies that
	\begin{align}\label{cor_MTL_error}
	%-\left[1- \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le 
	 & \left|L(\hat{\beta}_2^{\MTL}) -\frac{\sigma^2}{\rho_1+\rho_2-1} - pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\right|  \le  \left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c_{\varphi}}\sigma^2 ).
	 \end{align}
\end{corollary}

>>>>

\begin{corollary}[Derived from Theorem \ref{thm_many_tasks}]\label{prop_same_cov}
	In the setting described above, if the sample size $n$ is greater than $(1 + \tau) p$, then the total prediction loss of the $t$ tasks satisfies that
	\begin{align*} 
	 \sum_{i=1}^t L(\hat{\beta}_i^{\MTL})&= (t-r) \kappa^2 \cdot \frac{\tr\Sigma}{p}  + \sigma^2 \cdot \frac{r}{\rho-1} +\OO\left( p^{-c_\infty} \cdot\frac{\rho t(\kappa^2 + d^2) \|\Sigma\|}{(\sqrt{\rho}-1)^2 \kappa^2  \|\Sigma\| + \sigma^2}   \left( t \kappa^2 \|\Sigma \| + \frac{\sigma^2 r}{\rho-1} \right)  \right) .
	 %& = \sum_{i=r+1}^t \lambda_i^2 (\Sigma^{1/2} B^{\star}) + \sigma^2 \frac{r}{\rho-1} +\OO\left(  p^{-c_\infty} \frac{t(\kappa^2 + d^2)\cdot (p^{-1}\tr \Sigma)}{(\sqrt{\rho}-1)^2 \kappa^2 \cdot (p^{-1}\tr \Sigma) + \sigma^2}   \left( t \kappa^2 \frac{\tr \Sigma}{p} + \frac{\sigma^2 r}{\rho-1} \right) \right) \\ 
%	 &= (t-r) \kappa^2 \cdot \frac{\tr\Sigma}{p}  + \sigma^2 \cdot \frac{r}{\rho-1} \\
%	 &+\OO\left( p^{-c_\infty}t(\kappa^2+d^2) \cdot p^{-1/2} \|\Sigma\|_F + p^{-c_\infty} \frac{t(\kappa^2 + d^2)\cdot (p^{-1}\tr \Sigma)}{(\sqrt{\rho}-1)^2 \kappa^2 \cdot (p^{-1}\tr \Sigma) + \sigma^2}   \left( t \kappa^2 \frac{\tr \Sigma}{p} + \frac{\sigma^2 r}{\rho-1} \right)  \right) .
	\end{align*}
%and $\lambda_{r+1} (\Sigma^{1/2} B^{\star}) \ge \cdots\ge  \lambda_{t} (\Sigma^{1/2} B^{\star})$ denote the smallest $(t-r)$ singular values of $\Sigma^{1/2} B^{\star}$.
\end{corollary}
On the other hand, the total STL predication loss is 
\begin{align*}
 \sum_{i=1}^t L(\hat{\beta}_i^{\STL}) =\sigma^2\cdot \frac{t}{ \rho-1}  +\OO(p^{-c_\infty}\cdot t \sigma^2 ).
	 \end{align*}

@@

\begin{corollary}[Derived from Theorem \ref{thm_main_RMT}]\label{cor_MTL_loss}
	In the setting described above, assume further that a) the sample size $n_2$ of task two is greater than $(1 + \tau) p$, and b) there exists a small constant $c_0>0$ such that the signal-to-noise ratios satisfy 
	\be\label{choiceofpara0}
		p^{ c_0} \le \frac{p\kappa^2}{\sigma^2 }  \le p^{1-c_0},\quad  \frac{pd^2}{\sigma^2}=\OO(1).
	\ee
	%Let $B, A_1, A_2$ be the global minimum of equation \eqref{eq_mtl}. Then
	The prediction loss of the hard parameter sharing estimator for task two satisfies that
	\begin{align}\label{cor_MTL_error}
	%-\left[1- \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le 
	 & \left|L(\hat{\beta}_2^{\MTL}) -\frac{\sigma^2}{\rho_1+\rho_2-1} - pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\right|  \le  \left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2 ),
	 \end{align}
	 for any constant $c\in (0,c_0)$.
\end{corollary}

### 2020-09-30 arxiv/proof_examples.tex
revise

Under the two-task setting of Corollary \ref{cor_MTL_loss}, the optimization objective \eqref{eq_mtl} becomes
	\begin{align}
		f(A, B) =   \bignorm{X^{(1)} B A_1 - Y^{(1)}}^2+ \bignorm{X^{(2)} B A_2 - Y^{(2)}}^2, \label{eq_mtl_2task_cov}
	\end{align}
	where $B \in \real^{p}$ and $A_1, A_2\in \R$ because $r=1$ in the two-task case. % for $1 < r < t$ by our assumption.
	Without loss of generality, we assume that both $A_1$ and $A_2$ are nonzero, since otherwise we can add a small perturbation $\delta$ to them and later remove this perturbation by taking $\delta\to 0$ in the final result. Again using the local optimality condition ${\partial f}/{\partial B} = 0$, for any $A_1$ and $A_2$ we obtain the optimal $\hat{B}$ as follows
	\begin{align}
		\hat{B} \define  (A_1^2 (X^{(1)})^{\top}X^{(1)} + A_2^2 (X^{(2)})^{\top}X^{(2)})^{-1} (A_1 (X^{(1))})^{\top}Y^{(1)} + A_2 (X^{(2)})^{\top}Y^{(2)}). \label{eq_Bhat_2task} %\\
		%&= (B^\star A ^{\top}) (A A^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{j=1}^t \varepsilon_i A_i^{\top}} (A  A^{\top})^{-1}.
	\end{align}
 Now inserting $\hat B$ back into \eqref{eq_mtl_2task_cov}, we obtain an objective that only depends on $A:=A_1/A_2$ as follows 
 \begin{align}
		g(A) \define  \sum_{i=1}^2 . \label{eq_mtl_A12}
	\end{align}

@@

Plugging $Y^{(1)}$ and $Y^{(2)}$ in \eqref{model_YvsX} into \eqref{eq_Bhat_2task}, we can rewrite that 
 \begin{align}
		\hat{B}= \frac{1}{A_2}\left[\beta_2 + \left(\frac{A_1^2}{A_2^2}(X^{(1)})^{\top}X^{(1)} +  (X^{(2)})^{\top}X^{(2)}\right)^{-1}\bigbrace{ (X^{(1)})^{\top}X_1\left(\frac{A_1}{A_2}\beta_1 - \frac{A_1^2}{A_2^2} \beta_2\right) + \left(\frac{A_1}{A_2} (X^{(1)})^{\top}\varepsilon_1 + (X^{(2)})^{\top}\varepsilon_2\right)}\right]. \label{eq_Bhat_2task2} %\\
		%&= (B^\star A ^{\top}) (A A^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{j=1}^t \varepsilon_i A_i^{\top}} (A  A^{\top})^{-1}.
	\end{align}
\todo{next derive the prediction loss} 
 \cob

 \begin{claim}
	In the setting described above, we have that
	\be\label{approxvalid}
		g(v)=  \left[\val(v) + (N_1\sigma^2_1+N_2\sigma^2_2)\right]\cdot \left( 1+\OO(p^{-(1-\e_0)/2+\e})\right).
	\ee
\end{claim}
\begin{proof}
	We use the fact that our random vectors have i.i.d. entries.
%Before doing that, we first need to fix the setting for the following discussions, because we want to keep track of the error rate carefully instead of obtaining an asymptotic result only.
	Recall that $Y_i = X_i\beta_i + \varepsilon_i$ and $\wt Y_i = \wt X_i\beta_i + \wt\varepsilon_i$, $i=1,2$, all satisfy Assumption \ref{assm_secA2}. Then we rewrite \eqref{eq_mtl_2tasktilde} as
$$	g( v) = \sum_{i=1}^2\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2 , \quad \wt\beta:=\hat B w_i-\beta_i.$$
Since $ \wt X_i\wt\beta$ and $ \wt \e_i$ are independent random vectors with i.i.d. centered entries, we can use the concentration result, Lemma \ref{largedeviation}, to get that for any constant $\e>0$,
\begin{align*}
\left|\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2 -  \exarg{\wt X_i,\wt{\e}_i} {\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2} \right| & =\left|\left\| \wt X_i\wt\beta_i  - \wt \e_i\right\|^2 - N_i (\wt\beta_i^\top \Sigma_i \wt\beta_i + \sigma_i^2) \right| \\
&\le N_i^{1/2+\e} (\wt\beta_i^\top \Sigma_i \wt\beta_i + \sigma_i^2),
\end{align*}
with high probability. Thus we obtain that
$$g(v)= \left[\sum_{i=1}^2 N_i\left\|\Sigma_i^{1/2}( \hat B w_i - \beta_i) \right\|^2 + (N_1\sigma^2_1+N_2\sigma^2_2)\right]\cdot \left( 1+\OO(p^{-(1-\e_0)/2+\e})\right),$$
where we also used $N_i\ge p^{-1+\e_0}$. Inserting \eqref{hatB} into the above expression and using
 again the concentration result, Lemma \ref{largedeviation}, we get that
$$ \sum_{i=1}^2 N_i\left\|\Sigma_i^{1/2}( \hat B w_1 - \beta_i) \right\|^2 = \val(v)\cdot \left( 1+\OO(p^{-1/2+\e})\right)$$
with high probability.
%-----old-------
%Suppose that the entries of $\e_1$ and $\e_2$ have variance $\sigma^2$.  Using a validation set that is sub-sampled from the original training dataset, we get a validation loss as follows
%\begin{align}
%		&\val(\hat{B}; w_1, w_2):= \exarg{\varepsilon_1,\e_2} \sum_{i=1}^2 \left\|\Sigma_i^{1/2}( \hat B w_1 - \beta_i) \right\|^2 \\
%	&=  n_1 \cdot \bignorm{\Sigma_1^{1/2}\left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}X_2^{\top}X_2\left (\beta_s - \frac{w_1}{w_2}\beta_t\right)}^2 \nonumber \\
%		&+ n_1 \sigma^2 \cdot \frac{w_1^2}{w_2^2} \bigtr{\left(\frac{w_1^2}{w_2^2}  X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\Sigma_1} \nonumber \\
%		&+ n_2 \cdot \frac{w_1^2}{w_2^2}\bignorm{\Sigma_2^{1/2}\left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}X_1^{\top}X_1\left(\beta_s - \frac{w_1}{w_2}\beta_t\right)}^2 \nonumber \\
%		&+ n_2 \sigma^2 \cdot \bigtr{\left(\frac{w_1^2}{w_2^2} X_1^{\top}X_1 + X_2^{\top}X_2\right)^{-1}\Sigma_2}. \label{eq_val_mtl}
%\end{align}
%\nc
%------------------
Thus we conclude the proof.
\end{proof}

\nc

>>>>

Under the two-task setting of Corollary \ref{cor_MTL_loss}, the optimization objective \eqref{eq_mtl} becomes
	\begin{align}
		f(A, B) =   \bignorm{X^{(1)} B A_1 - Y^{(1)}}^2+ \bignorm{X^{(2)} B A_2 - Y^{(2)}}^2, \label{eq_mtl_2task_cov}
	\end{align}
	where $B \in \real^{p}$ and $A_1, A_2\in \R$ because $r=1$ in the two-task case. % for $1 < r < t$ by our assumption.
	Without loss of generality, we assume that both $A_1$ and $A_2$ are nonzero, since otherwise we can add a small perturbation $\delta$ to them and later remove this perturbation by taking $\delta\to 0$ in the final result. Again using the local optimality condition ${\partial f}/{\partial B} = 0$, for any $A_1$ and $A_2$ we obtain the optimal $\hat{B}$ as follows
	\begin{align}
		\hat{B} \define  (A_1^2 (X^{(1)})^{\top}X^{(1)} + A_2^2 (X^{(2)})^{\top}X^{(2)})^{-1} (A_1 (X^{(1))})^{\top}Y^{(1)} + A_2 (X^{(2)})^{\top}Y^{(2)}). \label{eq_Bhat_2task} %\\
		%&= (B^\star A ^{\top}) (A A^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{j=1}^t \varepsilon_i A_i^{\top}} (A  A^{\top})^{-1}.
	\end{align}
 Now inserting $\hat B$ back into \eqref{eq_mtl_2task_cov}, we obtain an objective that only depends on $A:=A_1/A_2$ as follows 
 \begin{align}
		 g(A) \define & \left\| X^{(1)} \hat \Sigma^{-1} (X^{(2)})^\top X^{(2)} (A\beta_2-\beta_1)+ \left(A^2 X^{(1)}\hat \Sigma^{-1} (X^{(1)})^\top-\id_{n_1}\right)\epsilon^{(1)} + A X^{(1)}\hat \Sigma^{-1} (X^{(2)})^\top \epsilon^{(2)} \right\|^2 \nonumber\\
		 + & \left\| X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)} (A\beta_1-A^2\beta_2)+ \left(X^{(2)}\hat \Sigma^{-1} (X^{(2)})^\top-\id_{n_2}\right)\epsilon^{(2)} + A X^{(2)}\hat \Sigma^{-1} (X^{(1)})^\top \epsilon^{(1)} \right\|^2, 
		\label{eq_mtl_A12}
	\end{align}
	where we denote
	$$\hat \Sigma:= A^2 (X^{(1)})^\top X^{(1)}  + (X^{(2)})^\top X^{(2)} .$$
We can calculate the partial expectation of $g(A)$ over the randomness in $\epsilon_1$ and $\epsilon_2$:
&\exarg{\epsilon^{(1)}, \epsilon^{(2)}}{g(A) \mid X_1, X_2 ,\beta_1,\beta_2} \\
&= \left\| X^{(1)} \hat \Sigma^{-1} (X^{(2)})^\top X^{(2)} (A\beta_2-\beta_1)\right\|^2 +\left\| X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)} (A\beta_1-A^2\beta_2)\right\|^2  \\
& + \exarg{\epsilon^{(1)}}{(\epsilon^{(1)})^\top \left(A^2 X^{(1)}\hat \Sigma^{-1} (X^{(1)})^\top-\id_{n_1}\right)^2\epsilon_1} +\exarg{ \epsilon^{(2)}}{(\epsilon^{(2)})^\top \left(X^{(2)}\hat \Sigma^{-1} (X^{(2)})^\top-\id_{n_2}\right)^2\epsilon^{(2)}}   \\
&+  A^2 \exarg{ \epsilon^{(2)}}{ (\epsilon^{(2)} )^\top X^{(2)} \hat \Sigma^{-1}  (X^{(1)})^\top X^{(1)}\hat \Sigma^{-1} (X^{(2)})^\top \epsilon^{(2)}}  +A^2 \exarg{ \epsilon^{(1)}} { (\epsilon^{(1)})^\top X^{(1)} \hat \Sigma^{-1}  (X^{(2)})^\top X^{(2)}\hat \Sigma^{-1} (X^{(1)})^\top \epsilon^{(1)}  }\\
		&= \left\| X^{(1)} \hat \Sigma^{-1} (X^{(2)})^\top X^{(2)} (A\beta_2-\beta_1)\right\|^2 +\left\| X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)} (A\beta_1-A^2\beta_2)\right\|^2  \\
		&+ \sigma^2\tr\left[ \left(A^2 X^{(1)}\hat \Sigma^{-1} (X^{(1)})^\top-\id_{n_1}\right)^2\right] +\sigma^2\tr\left[ \left(X^{(2)}\hat \Sigma^{-1} (X^{(2)})^\top-\id_{n_2}\right)^2 \right]\\
&+ 2 A^2 \sigma^2 \tr\left[ \hat \Sigma^{-1}  (X^{(1)})^\top X^{(1)}\hat \Sigma^{-1} (X^{(2)})^\top X^{(2)} \right]  \nonumber\\
		& = \left\| X^{(1)} \hat \Sigma^{-1} (X^{(2)})^\top X^{(2)} (A\beta_2-\beta_1)\right\|^2 +\left\| X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)} (A\beta_1-A^2\beta_2)\right\|^2 +\sigma^2(n_1+n_2-p).
Using the concentration estimates in Lemma \ref{largedeviation}, we can control the difference between $g(A)$ and its partial expectation. %$\exarg{\epsilon^{(1)}, \epsilon^{(2)}}{g(A) \mid X_1, X_2 ,\beta_1,\beta_2}$.
 \begin{claim}\label{claim_largedev1}
 In the setting of Proposition \ref{cor_MTL_loss}, we have that 
$$\left| g(A) - \exarg{\epsilon^{(1)}, \epsilon^{(2)}}{g(A) \mid X_1, X_2 ,\beta_1,\beta_2}\right| \le p^{-c_\infty}\cdot p\left(\sigma^2 +\|\beta_1\|^2 +\|\beta_2\|^2 \right). $$
 \end{claim}
 \begin{proof}
 \todo{add proof}

For the random-effects model, we can further take the partial expectation over $\beta_1$ and $\beta_2$:

 \begin{claim}\label{claim_largedev2}
 In the setting of Proposition \ref{cor_MTL_loss}, we have that 
$$\left| g(A) - \exarg{\epsilon^{(1)}, \epsilon^{(2)},\beta_1,\beta_2}{g(A) \mid X_1, X_2 ,\beta_1,\beta_2}\right| \le p^{-c_\infty}\cdot p\left(\sigma^2 +\|\beta_1\|^2 +\|\beta_2\|^2 \right). $$
 \end{claim}
 \begin{proof}
 \todo{add proof}
 \end{proof}

@@

Plugging $Y^{(1)}$ and $Y^{(2)}$ in \eqref{model_YvsX} into \eqref{eq_Bhat_2task}, we can rewrite that 
 \begin{align}
		\hat{B}= \frac{1}{A_2}\left[\beta_2 + \left(\frac{A_1^2}{A_2^2}(X^{(1)})^{\top}X^{(1)} +  (X^{(2)})^{\top}X^{(2)}\right)^{-1}\bigbrace{ (X^{(1)})^{\top}X_1\left(\frac{A_1}{A_2}\beta_1 - \frac{A_1^2}{A_2^2} \beta_2\right) + \left(\frac{A_1}{A_2} (X^{(1)})^{\top}\varepsilon_1 + (X^{(2)})^{\top}\varepsilon_2\right)}\right]. \label{eq_Bhat_2task2} %\\
		%&= (B^\star A ^{\top}) (A A^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{j=1}^t \varepsilon_i A_i^{\top}} (A  A^{\top})^{-1}.
	\end{align}
\todo{next derive the prediction loss}
